Mergeable Types for Highly-Available Replicated Data

### 2017-02-24 paper/meta.tex
Meta

We will now present a series of definitions and results that let us
understand the premises of the \rulelabel{E-Pull-Wait} rule, examine
the merge operation in closer detail, and appreciate the need for
constraining the branching structure. First, we formalize the
intuitive notation of the ancestor relationship between versions of a
legal branching history (i.e., a branching history generated by the
rules in Fig.~\ref{fig:opsem}):

@@

Fortunately, unlike the source control systems where branching
structure is entirely dictated by the user, \name abstracts away the
branching structure from the programmer, hence retains the ability to
constrain it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs by suitably constraining the branching
structure such that the problem never arises. The constraints are
imposed either implicitly, as a result of how operational semantics
defines an atomic step, or explicitly, by insisting that certain
conditions be met before merging a pair of versions
(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
disables criss-cross merges since it only ever merges versions that
are latest on their respective branches. For instance let $v_{11}$ and
$v_{21}$ be the latest versions on branches $b_1$ and $b_2$,
respectively. Meging $b_2$ into $b_1$ requires entails $v_{21}$ into
$v_{11}$ to generate version $v_{12}$ on $b_1$. Now, merging $b_1$
into $b_2$ translates to merging $v_{12}$ into $v_{21}$ (a
\emph{fast-forward} merge, in Git parlance), but not $v_{11}$ into
$v_{21}$, thus preventing a criss-cross branching structure. In other
words, a criss-cross branching structure is prevented due to the
order among merges between conflicting branches introduced as a result
of \rulelabel{E-Pull-Wait} being an atomic step.

Secondly, we impose certain pre-conditions on the merging branches to
preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
intuition is as follows: consider the branch $b_1$ at the instance of
merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
($v_{22}$) could be a common ancestor for a version on $b_1$
($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
($v_{32}$) could be a common ancestor for the new version on $b_1$
($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
ancestor of $v_{13}$, and both ancestors are not ordered by the
ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
$v_{44}$. We observe that this scenario can be prevented if, when
merging $b_3$, $b_1$ insists on an ancestor relation between the
merged version ($v_{22}$) of the $b_2$, the previously merged branch,
and the latest version of $b_3$, the current merging branch. We call
the last merged version the \emph{external locus} of the branch. By
requiring that, for every branch $b$, the external locii of $b$ at
various points in time (i.e., external locus of every prefix of $b$)
be totally ordered, we effectively enforce the invariant that for any
two common ancestors $v_1$ and $v_2$ between two versions, there
exists another common ancestor $v_3$ that succeeds $v_1$ and $v_3$ in
the ancestor relation, thereby preventing the possibility of multiple
LCAs.

%% <false>
%% Multiple versions that merge any pair of versions are totally
%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
%% </false>

%% If $v_1$ and $v_2$ are ancestors of $v_3$ and $v_4$, then there
%% exists a $v_5$, a successor of $v_1$ and $v_2$, and an ancestor of
%% $v_3$ and $v_4$
%% If I follow one of your locii, or you follow one of my locii, then
%% we both have same set of external common ancestors.

\begin{figure}
\centering
\includegraphics[scale=0.75]{Figures/legal-merge}

\caption{A legal branch history for the example in
Fig.~\ref{fig:external-lcas}}
\label{fig:legal-merge}
\end{figure}

Let us see how the aforementioned restriction allows us to merge the
branches in Fig.~\ref{fig:external-lcas} without ever creating
multiple LCAs. The legal branching history is shown in
Fig.~\ref{fig:legal-merge}. First, like in
Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
relation. Same applies for $b_1$ and $b_4$ because neither one's
latest version follows from other's external locus.  However, $b_2$
and $b_3$ can merge among themselves. Let us say they indeed merge to
to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
eligible to be merged into $b_1$ and also $b_4$. These merges do occur
in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
$v_{44}$, and making $v_{23}$ the external locus for both $b_1$ and
$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{45}$
on $b_4$. Note that alternative legal histories are also possible. For
instance, $b_4$ could have skipped the merge with $v_{23}$ (hence, the
version $v_{44}$), and directly merged $v_{13}$ in the last step.
This is legal because $b_4$'s locus before the merge, $v_{32}$, is an
ancestor of the merging version $v_{13}$.

We now formalize the intuitions described above to precisely define
the notion of mergeability:

\begin{definition} [\bfseries Internal and External Ancestors]
Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
is an ancestor from the same branch $b$. An external ancestor of $v$
is an ancestor from a different branch $b'\neq b$. 
\end{definition}

\begin{definition} [\bfseries External Locus]
Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
$v$ is an external ancestor that is not an ancestor of any other
external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
\end{definition}

\begin{definition} [\bfseries Mergeability]
Given a history $H$, a version $v_1$ and a version $v_2$ that is not
an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
$\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
ancestor of $v_2$, or $v_2$'s external locus is an ancestor of $v_1$.
\end{definition}

Rule \rulelabel{E-Pull-Wait} of the reduction relation
(Fig.~\ref{fig:opsem}) is enabled only if the latest version ($v'$)
of thread $t'$'s branch is mergeable into the latest version ($v$) of
thread $t$'s branch (i.e., $\under{H}{v' \mbleto v}$). Merge computes
the latest common ancestor of $v$ and $v'$, which is 
guaranteed to be unique as per the following theorem\footnote{Proof
included in the appendix}:

\begin{theorem} [\bfseries Unique LCA]
Every pair of versions $v_1$ and $v_2$ in a legal branch history $H$ 
have a unique least common ancestor. 
\end{theorem}

>>>>

To define the semantics of LCA, we formalize the intuitive notation of
the ancestor relationship between versions of a legal branching
history (i.e., a branching history generated by the rules in
Fig.~\ref{fig:opsem}):

@@

Fortunately, unlike the source control systems where branching
structure is entirely dictated by the user, \name abstracts away the
branching structure from the programmer, hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs either by constraining the branching
stucture to preempt multiple LCAs, or by extending it such that
merging branches have a new unique LCA. The preemption is done for
criss-cross merges by the operational semantics, which only allows
merging versions that are latest on their respective branches. For
instance let $v_{11}$ and $v_{21}$ be the latest versions on branches
$b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
merging $v_{21}$ into $v_{11}$ to generate version $v_{12}$ on $b_1$.
Now, merging $b_1$ into $b_2$ involves merging $v_{12}$, the latest
version on $b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus
preempting a criss-cross branching structure. 
% In other words, a criss-cross branching structure is prevented due
% to the order among merges between conflicting branches introduced as
% a result of \rulelabel{E-Pull-Wait} being an atomic step.

Next, for the branching structures where a pair of branches (i.e.,
their respective latest versions) have multiple externally-located
LCAs, the problem can be fixed by suitably extending the branching
structure such that the branches have a new single LCA. For instance,
consider the example in Fig.~\ref{fig:external-lcas}, where (latest
versions on) branches $b_1$ and $b_4$ have two externally-located LCAs
($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
however be reclaimed by extending the branching structure as shown in
Fig.~\ref{fig:legal-extension-1}. The extension is done by first merging
$b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then merging
$b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and finally
merging $b_2$ into $b_4$ to create version $v_{45}$ on $b_4$. Observe
that $b_1$ and $b_4$ now have a single LCA ($v_{23}$) hence are now
mergeable. Mergeability could also have been reclaimed by adopting an
alternative extension to the branching structure as shown in
Fig.~\ref{fig:legal-extension-2}.

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  $b_3$ is merged into $b_2$ to create $v_{23}$ here.
  \label{fig:legal-extension-1}
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/legal-extension-1}
}
\hfill
\subcaptionbox[] {\small
  $b_2$ is merged into $b_3$ to create $v_{33}$ here.
  \label{fig:legal-extension-2}
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/legal-extension-2}
}
\caption{Legal extensions of the branching structure in
  Fig.~\ref{fig:external-lcas} that reclaim the mergeability of $b_1$
  and $b_4$}
\label{fig:legal-extensions}
\end{figure}

Note that the extension strategy described above is possible only
because operational semantics imposes no obligation to merge any two
concurrent versions. The high-level purpose of merging branches in
\name is to propagate the data across the branches, which can be done
by merging any two concurrent versions that contain the data; the
semantics lets us pick these versions. While mergeability of branches
with multiple LCAs can be reclaimed by actively extending the
branching structure as described above, the semantics is
\emph{self-correcting} in the sense that a fair execution reclaims the
mergeability of any two branches without any intervention. This is
possible because the system preempts criss-cross merges, thus letting
the semantics make progress from any state. The property is formalized
by the following theorem\footnote{The more precise statement of the
theorem, and its proof are included in the appendix}:

\begin{theorem} [\bfseries Semantics Self-Correcting]
??
\end{theorem}

### 2017-02-24 paper/semantics.tex
Meta

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The premises of the
\rulelabel{E-Pull-Wait} serve the important purpose of contraining the
branching structure by allowing only the legal merges, and
consequently preserving certain desirable properties of the system;
we will go into the details shortly. The \rulelabel{E-Pull-Wait} and
\rulelabel{E-Pull} rules thus let a thread sync with a subset of
concurrent threads in multiple steps before returning the result of
the \C{pull}. Since \C{sync} is a composition of \C{push} and
\C{pull}, its behavior can be explained thus: \C{sync} pushes the
given value onto the current (local) branch, merges a (possibly empty)
subset of concurrent branches into the local branch, and returns the
result.

>>>>

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{least common ancestor} (LCA) for the
latest versions on the given pair of branches. The semantics of LCA is
discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
rules thus let a thread sync with a subset of concurrent threads in
multiple steps before returning the result of the \C{pull}. Since
\C{sync} is a composition of \C{push} and \C{pull}, its behavior can
be explained thus: \C{sync} pushes the given value onto the current
(local) branch, merges a (possibly empty) subset of concurrent
branches into the local branch, and returns the result.
